---
title: "Phenol Red - pH Indicator"
author: "Glenn Davis"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
bibliography: bibliography.bib
csl: personal.csl    
vignette: >
  %\VignetteIndexEntry{Phenol Red - pH Indicator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Phenol red is an indicator commonly used to measure pH
in swimming pool test kits, see e.g. @K-1000.
The goal of this **colorSpec** vignette is to reproduce the colors seen in such
a test kit, for typical values of pool pH.
Calculations like this one might make a good project for a college freshman chemistry class.
Featured functions in this vignette are: `interpolate()`, `calibrate()`,
and `plotPatchesRGB()`.

```{r, echo=TRUE, message=FALSE}
library( colorSpec )
```

<br>

## Absorbance Spectra at Different pH Values

The absorbance data for phenol red has already been digitized from @Rovati:

```{r fig.width=7, fig.height=4,  fig.show='hold', dev='png' }
path = system.file( "extdata/stains/PhenolRed-Fig7.txt", package="colorSpec" )
wave = 350:650
phenolred = readSpectra( path, wavelength=wave )
par( omi=c(0,0,0,0), mai=c(0.6,0.7,0.4,0.2) )
plot( phenolred, main='Absorbance Spectra of Phenol Red at Different pH Values' )
```
Compare this plot with @Rovati, Fig. 7.
Unfortunately, the concentration and optical path length are unknown,
but these curves can still be used as 'relative absorbance'.


<br>

## Absorbance at Selected Wavelengths

We investigate how absorbance depends on pH for a few selected wavelengths.

```{r, echo=TRUE, message=TRUE, fig.width=7, fig.height=5,  fig.show='hold' , dev='png' }
wavesel = c( 365, 430, 477, 520, 560, 590 )  # 365 and 477 are 'isosbestic points'
mat = apply( as.matrix(wavesel), 1, function( lambda ) { as.numeric(lambda == wave) } )
colnames( mat ) = sprintf( "%g nm", wavesel )
mono = colorSpec( mat, wavelength=wave, quantity='power' )
RGB = product( mono, BT.709.RGB, wavelength=wave )
colvec = grDevices::rgb( DisplayRGBfromLinearRGB( RGB / max(RGB) ) )

phenolsel = resample( phenolred, wavesel )
pH = as.numeric( sub( '[^0-9]*([0-9]+)$', '\\1', specnames(phenolred) ) )
pHvec = seq(min(pH),max(pH),by=0.05)
phenolsel = interpolate( phenolsel, pH, pHvec )
mat = t( as.matrix( phenolsel ) )
par( omi=c(0,0,0,0), mai=c(0.8,0.9,0.6,0.4) )
plot( range(pH), range(mat), las=1, xlab='pH', ylab='absorbance', type='n' )
grid( lty=1 ) ; abline( h=0 )
matlines( pHvec, mat, lwd=3, col=colvec, lty=1 )
title( "Absorbance of Phenol Red at Selected Wavelengths")
legend( 'topleft', specnames(mono), col=colvec, lty=1, lwd=3, bty='n' )
```

<br>

Note that the curves for the isosbestic points 365 and 477 nm are approximately flat,
as expected.
But for 430 nm the curve is distinctly non-monotone.
This indicates that the solution is not truly a mixture of the
acidic and basic species (especially for pH $\le$ 6),
and there may be an undesired side reaction, see @wiki:pH.

<br>

## Interpolation from pH=6.8 to pH=8.2

Swimming pools should be slightly basic;
a standard test kit covers the range from pH=6.8 to pH=8.2.
```{r, echo=TRUE,  message=TRUE, results='hold', fig.width=7, fig.height=4,  fig.show='hold'}
pHvec = seq(6.8,8.2,by=0.2)
phenolpool = interpolate( phenolred, pH, pHvec )
par( omi=c(0,0,0,0), mai=c(0.6,0.7,0.4,0.2) )
plot( phenolpool, main="Absorbance Spectra of Phenol Red at Swimming Pool pH Values" )
```

The rest of this section is best viewed on a display calibrated for sRGB, see @wiki:sRGB.
```{r, echo=TRUE,  message=TRUE }
# create an uncalibrated 'material responder'
testkit = product( D65.1nm, 'solution', BT.709.RGB, wave=wave )
# now calibrate so that fully transparent pure water has response RGB=c(1,1,1)
testkit = calibrate( testkit, response=1 )
RGB = product( phenolpool, testkit )
RGB
```
Unfortunately, in some cases the red value is greater than 1 (G and B are OK).
The color is outside the sRGB gamut.
Start over and recalibrate.
```{r, echo=TRUE,  message=TRUE }
testkit = product( D65.1nm, 'solution', BT.709.RGB, wave=wave )
# recalibrate, but lower the background a little, to allow more 'headroom' for indicator colors
bglin = 0.96  #  graylevel for the background, linear
testkit = calibrate( testkit, response=bglin )
RGB = product( phenolpool, testkit )
RGB
```
All values have been multiplied by `bglin`, and are now OK.
Draw the RGB patches on a white background multiplied by the same amount.
```{r, echo=TRUE, fig.width=7, fig.height=2.5,  fig.show='hold'}
class(RGB) = 'model.matrix'
df.RGB = data.frame( RGB=RGB, LEFT=1:nrow(RGB), TOP=0, WIDTH=1, HEIGHT=2 )
par( omi=c(0,0,0,0), mai=c(0.3,0,0.3,0) )
plotPatchesRGB( df.RGB, labels=F, background=bglin )
text( (1:nrow(RGB)) + 0.5, 2, sprintf("%.1f",pHvec), adj=c(0.5,1.2), xpd=NA )
title( main='Calculated Colors for pH from 6.8 to 8.2' )
```
The background color is that of pure water, and is not the full RGB=(255,255,255).

In the first figure above,
the phenol red concentration and optical path length are unknown.
Compared to a real test kit, the calculated colors look a little faded.
An absorbance multiplier can easily tweak the unknown concentration, as follows.

```{r, echo=TRUE, fig.width=7, fig.height=2.5,  fig.show='hold'}
tweak = 1.2
phenolpool = multiply( phenolpool, tweak )
RGB = product( phenolpool, testkit )
class(RGB) = 'model.matrix'
df.RGB = data.frame( RGB=RGB, LEFT=1:nrow(RGB), TOP=0, WIDTH=1, HEIGHT=2 )
par( omi=c(0,0,0,0), mai=c(0.3,0,0.3,0) )
plotPatchesRGB( df.RGB, background=bglin, labels=F )
text( (1:nrow(RGB)) + 0.5, 2, sprintf("%.1f",pHvec), adj=c(0.5,1.2), xpd=NA )
main = sprintf( 'Calculated Colors for pH from 6.8 to 8.2 (absorbance multiplier=%g)', tweak )
title( main=main )
```
These colors are a better match to those in the test kit.



## References

<div id="refs"></div>

<br><br>

## Session Information

<pre>
```{r, echo=FALSE, results='asis'}
sessionInfo()
```
</pre>
